---
title: "Shakespeare SOMs"
author: Daniel Tanner
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
   toc: True
   toc_depth: 2
   toc_float: True
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, fig.align='center', message = FALSE, warning = FALSE)
```

#GET Shakespeare DATA

##get play index table
```{r}
library("RCurl")
play_table <- read.csv(text=getURL("https://raw.githubusercontent.com/clarkdatalabs/soms/master/Shakespeare_txt/play_table.csv"), sep=",", header=T, check.names = FALSE)
```

##Read in play text
```{r}
library("readr")
for (row in row.names(play_table)){
  play_table[row, "text"] <- read_file(as.character(play_table[row,"path"]) , locale(encoding = "UTF-16"))
}
remove(row)
```

##Clean out bracketed tags

```{r}
play_table$text <- gsub('<[^>]*>','', play_table$text)
```

#Form & Clean Corpus

Form into corpus:
```{r form corpus}
library("tm")

play.vec <- VectorSource(play_table[,"text"])
play.corpus <- Corpus(play.vec)
```

Next we want to really clean up our texts. The tm package has functions for stripping punctuation, numbers, whitespace, the most common words (we don't want to be comparing counts of "and" and "to"), and converting everything to lowercase. We'll also use a function from the `SnowballC` package to cut words down to common stems for comparison.

_Note: this may take a few minutes to run._
```{r clean corpus 1}
library("SnowballC")

play.corpus <- tm_map(play.corpus, removePunctuation)
play.corpus <- tm_map(play.corpus, removeNumbers)
play.corpus <- tm_map(play.corpus, tolower)
play.corpus <- tm_map(play.corpus, removeWords, stopwords("english"))
play.corpus <- tm_map(play.corpus, stemDocument)
play.corpus <- tm_map(play.corpus, stripWhitespace)
play.corpus <- tm_map(play.corpus, PlainTextDocument)
play.corpus <- tm_map(play.corpus, removePunctuation)
```

Let's check out what some of the cleaned text looks like:
```{r inspect cleaned text}
substr(inspect(play.corpus[12]),1,500)
```

The functions that rely on english dictionaries had a bit of trouble with the Shakespearean English, but they're close enough.

##Document Term Matrix
Add in metadata for plotting

```{r make document term matrix}
play.DTM <- DocumentTermMatrix(play.corpus)
```

##Restrict to common words
so matrix not super sparse

```{r}
play.DTM.modified <- removeSparseTerms(play.DTM, 0.8)
```


##Normalize rows
- make each a length 1 vector
-add attributes for plotting
```{r}
play.DTM.modified <- t(apply(play.DTM.modified, 1, function(x) x / sqrt(sum(x^2)) ))
attr(play.DTM.modified, "name") <- play_table$name
attr(play.DTM.modified, "genre") <- play_table$genre

```



#SOM!

```{r}
library(kohonen)

play.SOM <- som(scale(play.DTM.modified), grid = somgrid(6,6, "hexagonal"),  toroidal = FALSE)
plot(play.SOM)
```

```{r}
plot(play.SOM, type = "mapping", col = palette(rainbow(3,v=.9))[as.integer(attr(play.DTM.modified, "genre"))], bgcol = "lightgray", labels = attr(play.DTM.modified, "name"))
```


This is not a particularly clean or useful visualization. There are two main problems that we will address below.
* There are too many labeled points, which become illegible.
* The gray backgrounds of teh cells don't communicate anything. It would be nice if they could be used to convey distance of a cell to it's negihbors, like in the `dist.neighbours` type SOMs.

First, we'll just make a list of the plays we want labeled, and represent the rest with *x*.

```{r}
plays.to.plot <- c("Romeo And Juliet", "Julius Caesar", "King Lear", "Macbeth", "Othello, the Moor of Venice", "The Tragedy of King Richard II", "The Tragedy of King Richard III", "The Tempest", "The Taming of the Shrew")
play.labels <- c()
X <- as.vector(attr(play.DTM.modified, "name"))
for (i in 1:37){
  if (X[i] %in% plays.to.plot){play.labels[i] <- X[i]}
  else{play.labels[i]<-"x"}
}
```

Next, we want to essentially combine two types of kohonen SOM functions, `mapping` and `dist.neighbours`. TO do this we will write our own function to calculate the distance of a cell's representative vector to all others, strongly weighing the neighboring cells. We'll then use this do set the background color using the `mapping` type SOM.

```{r}
code.distances <- function(SOM){
  N <- nrow(SOM$codes)
  #calculate all pairwise distances
  pairwise.distances = c()
  for (i in 1:N){
    list <- matrix()
    for (j in 1:N){
      list[j] <- dist(rbind(SOM$codes[i,],SOM$codes[j,]))
    }
    pairwise.distances <- cbind(pairwise.distances, list)
  }
  
  ##distance coefficient function, to be applied to geometric GRID distances
  d.coef<-function(x){
    if (x == 0){return(0)}
    else{return(1/(x^4))}
    }
  distance.coefficients <- apply(kohonen::unit.distances(SOM$grid, SOM$toroidal),c(1,2),d.coef)
  
  #calculate scaled sum of distances
  A<-distance.coefficients*pairwise.distances
  scaled.dist.sums <- colSums(A)/max(colSums(A))
  
  return(scaled.dist.sums)
}
```

Putting it all together:

```{r}
plot(play.SOM, type = "mapping", col = palette(rainbow(3,v=.9))[as.integer(attr(play.DTM.modified, "genre"))], bgcol = hsv(h=0,s=0,v=code.distances(play.SOM)), labels = play.labels)
```



#Character SOM

```{r}
library("RCurl")
character_table <- read.csv(text=getURL("https://raw.githubusercontent.com/clarkdatalabs/soms/master/Shakespeare_txt/character_table.csv"), sep=",", header=T, check.names = FALSE, stringsAsFactors = FALSE)
```

#DELETE Characters with AE in NAME!!!!
FIX THIS LATER!!

```{r}
#character_table <- character_table[-c(238,416,543,654,938,962,988,1089,1323),]
```



##Read in character lines

There are some errors accessing the text files with unicode characters in their names. I'll just throw some error handling around this and drop these characters from consideration.

```{r}
library("readr")
for (row in row.names(character_table)){
  tryCatch({
    character_table[row, "text"] <- read_file(as.character(character_table[row,"path"]) , locale(encoding = "UTF-16"))
  }, error=function(e){print(row)})
}
remove(row)
```

```{r}
character_table <- character_table[!is.na(character_table$text),]
```


Let's look at the metadata we have for each text file of lines:

```{r}
head(character_table[,c("character", "play", "genre")], 10)
```

Note the [*dramatis_personae*](https://en.wikipedia.org/wiki/Dramatis_person%C3%A6) character. Each play has one of these documents, which is just a list of characters and contains no lines. Let's get rid of them.

```{r}
character_table <- subset(character_table, character!="dramatis_personae")
```


The text here has the same bracketed tags, white space, punctuation, and uncommon words that we stripped out of the play text above, so we follow the same procedure to simplify it.

##Clean out bracketed tags

```{r}
character_table$text <- gsub('<[^>]*>','', character_table$text)
```

#Form & Clean Corpus

Form into corpus:
```{r form corpus}
library("tm")
character.corpus <- Corpus(VectorSource(character_table[,"text"]))
```

_Note: this may take a few minutes to run._
```{r clean character corpus}
library("SnowballC")

character.corpus <- tm_map(character.corpus, removePunctuation)
character.corpus <- tm_map(character.corpus, removeNumbers)
character.corpus <- tm_map(character.corpus, tolower)
character.corpus <- tm_map(character.corpus, removeWords, stopwords("english"))
character.corpus <- tm_map(character.corpus, stemDocument)
character.corpus <- tm_map(character.corpus, stripWhitespace)
character.corpus <- tm_map(character.corpus, PlainTextDocument)
character.corpus <- tm_map(character.corpus, removePunctuation)
```



##Doc Term Matrix

```{r}
character.DTM <- DocumentTermMatrix(character.corpus)
```


##Reduce Sparseness

```{r}
character.DTM.modified <- removeSparseTerms(character.DTM, 0.8)
```

##Normalize

```{r}
character.DTM.modified <- t(apply(character.DTM.modified, 1, function(x) x/sqrt(sum(x^2))))
attr(character.DTM.modified, "character") <- character_table$character
attr(character.DTM.modified, "genre") <- character_table$genre
```

strip NANs -----BLAH
```{r}
#####X <- character.DTM.modified[!is.na(character.DTM.modified[,1]),]
```


##PLOT!

First we make our SOM
```{r}
library(kohonen)
character.SOM <- som(scale(character.DTM.modified), grid = somgrid(30, 30, "hexagonal"), toroidal = FALSE)
```

```{r}
characters.to.plot <- c("Romeo", "Julius Caesar", "King Lear", "Macbeth", 
    "Othello", "Puck")
character.labels <- c()
X <- as.vector(attr(character.DTM.modified, "character"))
for (i in 1:length(X)) {
    if (X[i] %in% characters.to.plot) {
        character.labels[i] <- X[i]
    } else {
        character.labels[i] <- "x"
    }
}

```


